---
title: "taxonAlign"
format: gfm
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
<!-- badges: end -->

`taxonAlign` helps you:

1. **Build a personalised taxonomic reference** by combining one or more taxonomic reference
   tables -- your own, one built from the GBIF backbone taxonomy with
   `generate_GBIF_taxonomic_reference_list()`, a known source like the Australian Faunal Directory
   or the Australian Plant Census via `load_taxonomic_resources()`, or any mix of these.
2. **Fuzzy-match a list of raw taxon names** against that reference, resolving misspellings,
   synonyms, and names only identifiable to a higher rank (`genus sp.`), to maximise how many of
   your names align successfully.
3. **Update outdated names to the currently accepted/valid name** as long as the taxonomic reference 
   you are using includes synonyms and specifies the taxonomic status of all names.

It's inspired by [APCalign](https://traitecoevo.github.io/APCalign/)'s approach to aligning taxon
names to the Australian Plant Census, generalised so the reference doesn't have to be APC/APNI and
the taxa don't have to be plants and names of any taxonomic rank present in your reference (species, genus,
family, or anything else) can be matched.

## Installation

`taxonAlign` isn't on CRAN. Install the development version from GitHub:

``` r
# install.packages("pak")
pak::pak("traitecoevo/taxonAlign")
```

## Workflow

| Function | What it does |
|---|---|
| `generate_GBIF_taxonomic_reference_list()` | Build a reference table from the GBIF backbone taxonomy for a taxon (optionally filtered by country and/or minimum rank). |
| `load_taxonomic_resources()` | Fetch/reshape a known reference dataset into the same flat schema -- currently the Australian Faunal Directory (`"AFD"`) and the Australian Plant Census (`"APC"`, via `APCalign`). |
| `prepare_taxonomic_resources()` | Combine one or more reference tables (yours, GBIF's, a known source, or any mix) into the structure the matching functions need. |
| `align_taxa()` | Match a list of raw taxon names against a prepared reference, via exact and fuzzy matching. |
| `update_taxa()` | Resolve a matched (possibly synonymous) name forward to its current accepted name. |
| `create_taxonomic_update_lookup()` | Run `align_taxa()` and `update_taxa()` together in one call -- the easiest way to go from raw names to a lookup table of accepted names. |

## Example

First generated your taxonomic reference list, in this case, use `generate_GBIF_taxonomic_reference_list()` to fetch every name GBIF's backbone taxonomy has ever used for the painted lady butterfly, *Vanessa cardui*: its current accepted
name, plus dozens of historical synonyms (it's been reclassified several times, including a long stint
in the genus *Cynthia*).

```{r example}
library(taxonAlign)

gbif_reference <- generate_GBIF_taxonomic_reference_list("Vanessa cardui")
resources <- prepare_taxonomic_resources(gbif_reference)
```

`create_taxonomic_update_lookup()` then matches a list of raw names -- an already-correct name, a
genuine historical synonym, and a misspelling -- and resolves each to its current accepted name in one
call:

```{r}
create_taxonomic_update_lookup(
  c("Vanessa cardui", "Cynthia cardui", "Vanessa carduii"),
  resources
) |>
  dplyr::select(original_name, aligned_name, accepted_name, taxonomic_status_aligned)
```

`Cynthia cardui` -- the name this species was described under updates to the current accepted name, `taxonomic_status_aligned` records that it was a synonym, and the
misspelling (`carduii`, one letter off) is corrected via fuzzy matching.

For a known source instead of a live GBIF fetch, `load_taxonomic_resources()` fetches/reshapes a fixed
dataset -- the Australian Faunal Directory or the Australian Plant Census -- into the same schema:

```{r}
#| eval: false
afd <- load_taxonomic_resources("AFD")  # reads/reshapes a local AFD export -- see ?load_taxonomic_resources
apc <- load_taxonomic_resources("APC")  # wraps APCalign::load_taxonomic_resources()

# combine a known source with your own data, or two known sources together
resources <- prepare_taxonomic_resources(list(afd = afd$AFD, mine = my_reference))
```

See [`vignettes/get-started.qmd`](vignettes/get-started.qmd) for a full walkthrough of every function,
including building your own reference table by hand, combining several sources, matching names of
every rank (not just species), and an end-to-end "update a list of raw field names" workflow.

<!-- You'll still need to render `README.qmd` regularly, to keep `README.md` up-to-date.  -->
