---
title: "How and why taxonAlign was built"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  eval: false
  warning: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This document is not a tutorial -- [`get-started.qmd`](get-started.qmd) is that. It's a record of *why*
`taxonAlign` exists, the real workflow it was built to replace, and the real, messy invertebrate names
that shaped its design. Every number and every example name below is real, drawn from actually running
this package against actual data during its development -- not invented for illustration.

Every code chunk here is shown for the record rather than executed when this document renders: they
depend on files that aren't part of the package itself -- the ~89MB Australian Faunal Directory export,
a personal iNaturalist export, and the AusInvertTraits project's own name list -- so they can't run for
every reader the way [`get-started.qmd`](get-started.qmd)'s self-contained examples do. Revisits
[issue #2](https://github.com/traitecoevo/taxonAlign/issues/2), which originally asked for a benchmark
vignette reproducing the pre-package workflow; this is that idea, reimplemented against the finished
package rather than the prototype script it replaced.

## Why this package exists

[AusInvertTraits](https://github.com/traitecoevo/ausinvertraits) is a database of Australian
invertebrate trait data, assembled from hundreds of source datasets. Every one of those datasets
identifies its specimens in its own way -- and invertebrate taxonomy, in practice, is far messier than
"Genus species". A single AusInvertTraits source can contain, alongside ordinary binomials, names like:

- `"Genus 1 sp.01 Corinnidae"` -- a museum voucher/morphospecies code: unidentified to genus, with the
  family recorded separately.
- `"Aderid BF05"` -- an informal, vernacular-adjective family reference (family Aderidae) plus a
  collector's voucher code.
- `"Lasioglossum (Parasphecodes)"` -- the zoological `Genus (Subgenus)` bracket convention, rarely
  written by taxonomists working in plants (where `taxonAlign`'s design lineage, APCalign, comes from).
- `"Notonomus violaceus-grp. (BF01)"` -- a species group, not a single species.
- `"Themognatha affinis affinis"` -- a tautonymous subspecies (the nominotypical subspecies repeats its
  species epithet), textually indistinguishable from a hedge like `"Acacia aff. dealbata"` ("resembles
  *A. dealbata*, not confirmed").

None of these are edge cases dreamed up to stress-test a matching engine -- every one of them is a real
name that appeared in real AusInvertTraits source data. Aligning tens of thousands of names like these
against an accepted taxonomy, by hand, doesn't scale. The tool originally built to do this,
`AusInvertAlign` (in the sibling `ausinvertraits.addons` repository), was a single, long script: it read
one hard-coded combination of taxonomic references (the Australian Faunal Directory, AFD, supplemented
by an iNaturalist export for names AFD didn't cover) and ran a fixed sequence of match attempts against
them.

`taxonAlign` is that script's ideas, generalised into a reusable package: the reference is supplied by
the user (their own table, a known source, or a live GBIF fetch -- any mix of these), and the matching
engine handles *every* taxonomic rank present in that reference, not just genus/species/family the way
APCalign (`taxonAlign`'s direct design inspiration, built for the Australian Plant Census) does. The
original, still-unwrapped prototype script this was extracted from is kept for reference at
`ignore/match_taxa_for_inverts_202500304-use this.R`.

## Building the reference AusInvertAlign used: AFD + iNaturalist

`load_taxonomic_resources("AFD")` reads and reshapes the real, ~117,000-row AFD export shipped with the
package (`inst/extdata/AFD.csv`) into `taxonAlign`'s flat schema -- expanding into accepted
species/subspecies rows, one row per distinct higher-rank value (genus through phylum), and one row per
synonym (AFD embeds these as free-text, author-and-year-suffixed strings that have to be split and
stripped).

```{r}
library(taxonAlign)

afd <- load_taxonomic_resources("AFD")$AFD
nrow(afd)
#> [1] 310050
```

AFD alone doesn't cover every name AusInvertTraits needs -- some taxa (or their morphospecies-code
parent genera/families) are only in less formal, more actively-updated sources. `AusInvertAlign`
supplemented AFD with an export of iNaturalist's Australian invertebrate taxonomy. `taxonAlign` has no
built-in iNaturalist loader (unlike AFD and APC, iNaturalist has no single canonical export shape worth
building a permanent loader around), so this reshaping step is written out directly -- exactly the kind
of column-mapping `prepare_taxonomic_resources(interactive = TRUE)` exists to walk a user through
conversationally, done here as plain code instead:

```{r}
inat_raw <- readr::read_csv("inat_inverts_in_Australia.csv")

inat <- inat_raw |>
  dplyr::transmute(
    canonical_name = scientificName, scientific_name = scientificName, taxon_rank = taxonRank,
    taxonomic_status = "accepted", taxonomic_dataset = "iNat", genus = genus,
    taxon_ID = taxonID, accepted_name_usage_ID = taxonID
  )
nrow(inat)
#> [1] 54926
```

`prepare_taxonomic_resources()` combines the two into the single, rank-split, priority-ordered structure
`align_taxa()`/`update_taxa()` expect. Listing AFD first in the combined list means AFD wins whenever a
name is present in both -- the same priority AusInvertAlign gave it:

```{r}
resources <- prepare_taxonomic_resources(list(AFD = afd, iNat = inat))
names(resources)
#> [1] "species" "genus" "subgenus" "subgenus_v2" "complex" "hybrid" ... "family" ... "order" ...
```

## Aligning the real AusInvertTraits name list

AusInvertTraits keeps its own record of every raw name it has ever needed to align, and what it was
resolved to (`taxon_name`, filled in only when resolution reached species level -- a name resolved to a
higher rank leaves this blank, an "exception flag" rather than a failure). At the time of writing, this
list holds 5,154 distinct (name, source dataset) pairs.

```{r}
old <- readr::read_csv("AusInvertTraits_taxonomic_updates.csv")
nrow(old)
#> [1] 5154

updated <- create_taxonomic_update_lookup(
  old$original_name, resources,
  identifier = old$dataset_id,
  hybrids = TRUE, intergrades_affinis = TRUE, consider_english_name_endings = TRUE,
  include_bracketed_info = TRUE, progress = TRUE
)
```

`hybrids`/`intergrades_affinis` opt in to recognising the naming conventions above that can only ever
resolve to genus rank (a hybrid formula, an intergrade, a collector's indecision between two taxa, a
graded/"affinis"/"cf." identification); `consider_english_name_endings` opts in to recognising an
informal vernacular ending (`"Coccinellid"` for family Coccinellidae, `"Melolonthine"` for subfamily
Melolonthinae) and trying an exact match on the formal Latin equivalent before falling back to fuzzy
matching.

## How well does it work?

Comparing `updated$aligned_name` against the historical `taxon_name` directly under-states agreement,
because `taxon_name` is blank whenever the historical resolution reached a rank above species -- not
because the alignment failed. Restricting the comparison to the 2,873 rows where `taxon_name` is
actually filled in:

```{r}
comparison <- old |>
  dplyr::select(original_name, taxon_name) |>
  dplyr::bind_cols(updated |> dplyr::select(new_aligned_name = aligned_name)) |>
  dplyr::filter(!is.na(taxon_name), taxon_name != "")

nrow(comparison)
#> [1] 2873
mean(comparison$new_aligned_name == comparison$taxon_name, na.rm = TRUE)
#> [1] 0.9958... -- 2,861 of 2,873 agree
```

Only 12 names disagree with over a decade of manual curation, and looking at each one individually is
worth the effort -- not glossing over them, since they turn out to be four genuinely different kinds of
disagreement, only one of which was actually a package bug:

- **Six are species genuinely absent from both AFD and iNaturalist** (`"Acizzia errabunda"`,
  `"Castiarina lisaejessicae"`, `"Ephippiger ephippiger"`, `"Euastacus coughrani"`,
  `"Eurhopalus riguus"`, and `"Agrilus (Agrilus) thylacinus"` -- the last one also disagreeing on
  subgenus, AFD listing `Agrilus`'s nominotypical subgenus where the historical curation used a
  different one, `Pinarinus`). The pipeline conservatively stops at genus rather than guessing, which
  is the right behaviour, but it is a real coverage gap in these two reference sources.
- **Two look like genuine fuzzy-matching imprecision, not a coverage gap** -- `"Euastacus maccai"`
  fuzzy-matched to the real, but almost certainly wrong, `"Euastacus maidae"`, and
  `"Euastacus serratus"` fuzzy-matched to `"Euastacus sulcatus"` where the historical curation instead
  resolved it to `"Euastacus spinifer"`. *Euastacus* (the spiny freshwater crayfish) has dozens of
  species with very similarly-spelled epithets, and this is a real, honest limitation worth recording
  rather than glossing over: fuzzy matching within such a genus carries a real risk of landing on a
  plausible but wrong species, one this comparison caught but hasn't yet fixed
  ([issue #17](https://github.com/traitecoevo/taxonAlign/issues/17)).
- **Two are the same species, correctly matched identically both times** -- `"Onthophagus bulga"`
  appears twice in the historical file under two different source datasets, and the *historical* file
  itself records it inconsistently (`"Onthophagus [Onthophagus] bulga"` in one, `"Onthophagus bulga"`
  in the other) -- not a disagreement this package introduced.
- **One, `"Incertae Sedis"`**, resolves correctly to AFD's own literal "position uncertain" placeholder
  row rather than the more specific subfamily (`Megachilinae`) a human curator assigned from context a
  bare string can't carry -- a real limit of string-matching versus specimen-level knowledge, not a bug.
- **The last one, `"Themognatha affinis affinis"`, *was* a real bug**, and finding it is a good
  illustration of why this comparison is worth doing at all.

That tautonymous subspecies (mentioned above) was resolving only to genus rank, because
`APCalign::standardise_names()` abbreviates a bare `"affinis"` to `"aff."` regardless of whether it's a
genuine epithet or a hedge, making the two indistinguishable by text shape alone. Fixed in
[issue #16](https://github.com/traitecoevo/taxonAlign/issues/16) by teaching the matcher to check for a
real, resource-verified exact match before ever falling back to the "uncertain identification" reading
-- not by guessing that a repeated word always means a tautonym, which would just as easily have
upgraded a genuine hedge into a false, over-confident match in the other direction. With that fix,
agreement rises to 2,862 of 2,873 (99.6%).

## What real invertebrate names taught the matching engine

Several of `match_taxa()`'s ~20 sequential match blocks exist specifically because of patterns first
seen in this real data, not because they were anticipated in advance:

- **The bracketed subgenus convention** (`"Lasioglossum (Parasphecodes)"`, or the nominotypical form
  where subgenus repeats the genus, `"Aporocera (Aporocera) t-viride"`) needed its own dedicated match
  blocks -- a plain string-cleaning pass (borrowed from APCalign, which never has to handle this
  convention) strips the parentheses but keeps the bracketed word, which can otherwise be mistaken for a
  species epithet and fuzzy-matched to an unrelated real species (issue #14).
- **Morphospecies and voucher codes** (`"sp. 1"`, `"sp. nov."`, `"sp. indet."`, a bare collector's code)
  needed to resolve to the highest rank actually indicated, not fail outright or, worse, coincidentally
  fuzzy-match something unrelated -- and needed guarding against literal placeholder words ("Genus",
  "Family") that both GBIF's own undescribed-taxon convention *and* voucher codes independently use for
  "not yet identified", for entirely unrelated reasons (issue #13's rank-word fuzzy-match guard).
- **Fuzzy matching a higher rank needed to search broadest-first, not most-specific-first** -- checked
  directly against the real AusInvertTraits list: 52% of names that fuzzy-matched a higher rank *also*
  fuzzy-matched a real candidate at a different rank, and the large majority weren't coincidence but a
  systematic pattern -- an informal vernacular adjective (`"Melolonthine"`, `"Coccinellid"`) signals the
  broader group it derives from, not a specific genus (issue #12).
- **Hybrid, intergrade, indecision, and graded/"affinis" identifications** all needed to resolve to
  genus rank only, since none of them can specify a genuine species -- and the "affinis" case
  specifically needed to check for a real exact match first, for the tautonym reason above (issue #16).

None of this generalises to APCalign's own domain the way it does here, which is exactly the point: a
plant taxonomist's own name list essentially never contains a subgenus bracket, a hybrid marker, or a
BF-prefixed voucher code the way an invertebrate collection's does.

## Swapping the reference: AFD + iNaturalist vs. AFD + GBIF

`taxonAlign`'s reference is meant to be swappable. As a check that this is actually true in practice --
not just true in principle -- the same 5,154-name AusInvertTraits list was aligned a second time,
against AFD combined with a GBIF-backbone-derived reference of Australian-occurring invertebrates
instead of iNaturalist (see the next section for how that reference was built):

```{r}
gbif_au <- readRDS("gbif_au_invertebrates.rds")
resources_gbif <- prepare_taxonomic_resources(list(AFD = afd, GBIF = gbif_au))

updated_gbif <- create_taxonomic_update_lookup(
  old$original_name, resources_gbif,
  identifier = old$dataset_id,
  hybrids = TRUE, intergrades_affinis = TRUE, consider_english_name_endings = TRUE,
  include_bracketed_info = TRUE, progress = TRUE
)

mean(dplyr::coalesce(
  updated$aligned_name == updated_gbif$aligned_name,
  is.na(updated$aligned_name) & is.na(updated_gbif$aligned_name)
))
#> [1] 0.9913... -- 45 of 5,154 disagree
```

The two combinations agree on 99.1% of names. Where they disagree, it's overwhelmingly explained by
which resource happens to contain the specific name (a real species GBIF has that iNaturalist doesn't,
or vice versa) rather than a difference in how the matching engine itself behaves -- good evidence that
the reference and the matching engine really are two separable concerns, the way the package's design
intends.

## The GBIF chapter

`generate_GBIF_taxonomic_reference_list()` is the one piece of the package that doesn't need a
pre-existing export at all -- it builds a reference directly from the live GBIF backbone taxonomy for
any taxon:

```{r}
gbif_reference <- generate_GBIF_taxonomic_reference_list("Vanessa cardui")
```

That works well for a single genus or family. Building a reference for *every* Australian invertebrate
-- the scale AusInvertTraits actually needs -- turned out to be a substantially harder engineering
problem than fetching one small clade, and solving it is its own piece of this package's history:

- GBIF's taxonomy-browsing API refuses to page past 100,000 rows into any single clade (confirmed
  directly against the raw REST API, not just an `rgbif` quirk) -- a clade the size of Arthropoda (over
  three million descendants) has to be split recursively into smaller children and fetched piece by
  piece, with each piece cached independently so an interrupted multi-day fetch resumes for free rather
  than restarting.
- A country filter can't narrow the *fetch* itself -- GBIF's taxonomy-browsing endpoint has no country
  concept at all, only its occurrence-record endpoint does, and occurrence records are tagged with a
  taxon's accepted usage, essentially never a synonym's own key. The whole tree has to be fetched first
  and filtered to a country's occurrences afterward.
- Real network conditions at this scale -- transient TLS failures, requests that hang rather than error,
  occasional multi-minute API slowdowns under concurrent load -- needed real retry-with-backoff and
  timeout handling, tuned against actual measured behaviour rather than guessed at.

The result: a 358,496-row reference of GBIF-backed, Australian-occurring invertebrate taxa, and a
3,987,219-row worldwide version built the same way but without the country filter -- confirming, in the
process, that a single worldwide dataset built once could in principle serve *any* country's alignment
need, with the country filter applied locally and cheaply afterward, rather than refetched from scratch
per country. That reframing -- ship one worldwide reference, filter locally -- is a real architectural
option this package's design now supports, even though it isn't built as a released, downloadable
dataset yet.

## Lessons for whoever reads this next

A few things worth carrying forward, beyond the specific bugs and features above:

- **A hand-built test fixture finds the bugs you anticipated; a real, large, messy dataset finds the
  ones you didn't.** Several of the real bugs referenced above (issues #12, #13, #14, #16, #17) were only
  found by running the full, real AusInvertTraits list end to end and directly diffing the result
  against independent alignments or a real historical answer key -- not by unit tests alone, however
  thorough. The unit test fixtures in `tests/testthat/` exist to *keep* each of these fixed once found,
  not to have found them in the first place.
- **A plausible-sounding fix deserves the same scepticism as the bug it's fixing.** The first attempt at
  fixing issue #16 -- move the affinis/cf. check to run after ordinary exact species matching instead of
  guessing about repeated words -- was itself wrong, and only caught because a real, adversarial test
  (`test-apc_equivalence.R`, comparing against a live, independent implementation) existed to catch it.
- **A reference built for one purpose tends to be useful for another.** The GBIF invertebrate reference
  built to validate `taxonAlign` against AusInvertTraits is, unmodified, also a candidate worldwide
  taxonomic resource in its own right -- and the AFD/iNaturalist-vs-AFD/GBIF comparison above only
  existed because both references were already sitting on disk from other work.

See `CLAUDE.md` for the full technical detail behind every claim in this document, and the closed issues
on the [GitHub repository](https://github.com/traitecoevo/taxonAlign/issues?q=is%3Aissue) for the
discussion behind each fix.
