---
title: "How and why taxonAlign was built"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  eval: false
  warning: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This document is not a tutorial -- [`get-started.qmd`](get-started.qmd) is that. It's a record of *why*
`taxonAlign` exists and the real, messy invertebrate names that shaped its design. Every number and
example name below is real, drawn from actually running this package against real data during its
development.

The AFD, iNaturalist and GBIF reference exports this document uses are bundled with the package
(`inst/extdata/`), so most of the code below can genuinely be run, substituting in the invertebrate database's taxon list with one from your project.

## Why this package exists

An Australian invertebrates trait database assembled trait data from hundreds of source datasets, each
identifying its specimens in its own way -- a perfect example that invertebrate taxonomy is, in practice, far messier than
"Genus species". Real names from that database include things like:

- `"Genus 1 sp.01 Corinnidae"` -- a museum voucher/morphospecies code: unidentified to genus, with the
  family recorded separately.
- `"Lasioglossum (Parasphecodes)"` -- the zoological `Genus (Subgenus)` bracket convention, rarely
  written by taxonomists working in plants (where `taxonAlign`'s design lineage, APCalign, comes from).
- `"Themognatha affinis affinis"` -- a tautonymous subspecies (the nominotypical subspecies repeats its
  species epithet)

Aligning tens of thousands of names like these against an accepted taxonomy, by hand, doesn't scale.
The tool originally built to do this was a bespoke internal script, using ideas drawn from [APCalign](https://github.com/traitecoevo/APCalign): it read one hard-coded combination
of taxonomic references (the Australian Faunal Directory, AFD, supplemented by an iNaturalist export
for names AFD didn't cover) and ran a fixed sequence of match attempts against them.

`taxonAlign` generalises that script's ideas into a reusable package: the reference is supplied by the
user (their own table, a known source, or a live GBIF fetch -- any mix of these), and the matching
engine handles *every* taxonomic rank present in that reference, reflecting both the inability of researchers to identify some taxa to species -- or even genus -- and that some trait values are applicable across all species in a broader taxon grouping.

## Building the reference: AFD + iNaturalist

`load_taxonomic_resources("AFD")` reads and reshapes the real, ~117,000-row AFD export into
`taxonAlign`'s flat schema.

```{r}
library(taxonAlign)

afd <- load_taxonomic_resources("AFD")$AFD
nrow(afd)
#> [1] 310050
```

AFD's coverage isn't uniform across invertebrate phyla -- it's comprehensive for some well-studied
groups (insects especially) but considerably less complete for others, so it alone doesn't cover every
name a broad invertebrate database needs. The original workflow supplemented it with an export of
iNaturalist's Australian invertebrate taxonomy. `taxonAlign` has no built-in iNaturalist loader (unlike
AFD and APC, iNaturalist has no single canonical export shape worth building a permanent loader
around), so this reshaping step is written out directly:

```{r}
inat_raw <- readr::read_csv(system.file("extdata", "iNat_inverts_in_Australia.csv", package = "taxonAlign"))

inat <- inat_raw |>
  dplyr::transmute(
    canonical_name = scientificName, scientific_name = scientificName, taxon_rank = taxonRank,
    taxonomic_status = "accepted", taxonomic_dataset = "iNat", genus = genus,
    taxon_ID = taxonID, accepted_name_usage_ID = taxonID
  )
nrow(inat)
#> [1] 54926
```

Note `taxonomic_status` is hardcoded to `"accepted"` for every row -- this export carries no synonym
information at all, so iNaturalist can widen genus/species *coverage* beyond AFD but can never help
resolve a name that's a synonym of something else the way AFD's own synonym data can.

`prepare_taxonomic_resources()` combines the two into the single, rank-split, priority-ordered structure
`align_taxa()`/`update_taxa()` expect. Listing AFD first means it is listed as the taxonomic reference whenever a name is present in
both:

```{r}
resources <- prepare_taxonomic_resources(list(AFD = afd, iNat = inat))
names(resources)
#> [1] "species" "genus" "subgenus" "subgenus_v2" "complex" "hybrid" ... "family" ... "order" ...
```

## Aligning a real invertebrate name list

The Australian invertebrates database keeps its own record of every raw name it has ever needed to
align, and what it was resolved to -- in two separate columns. `taxon_name` records the species-level
resolution only, and is deliberately left blank whenever a name was correctly resolved to a *higher*
rank instead ("exception flag" rather than a failure). `aligned_name` records the resolution at
*every* rank, including the higher-rank cases `taxon_name` leaves blank, in the same bracketed
`"Genus sp. [original; identifier]"` format used throughout this package. Both are used below --
`taxon_name` for the species-level comparison, `aligned_name` for the higher-rank one. At the time of
writing, this list holds 5,154 distinct (name, source dataset) pairs.

```{r}
old <- readr::read_csv("your_own_copy_of_the_name_list.csv")
nrow(old)
#> [1] 5154

updated <- create_taxonomic_update_lookup(
  old$original_name, resources,
  identifier = old$dataset_id, # source dataset becomes identified in []
  hybrids = TRUE, 
  intergrades_affinis = TRUE,
  consider_english_name_endings = TRUE, # automatically replace English name endings with Latin ones
  include_bracketed_info = TRUE, # retain notes and original names in []
  progress = TRUE # toolbar to show progress
)
```

`hybrids`/`intergrades_affinis` opt in to recognising naming conventions that can only ever resolve to
genus rank (a hybrid formula, an intergrade, a collector's indecision between two taxa, a graded/
"affinis"/"cf." identification); `consider_english_name_endings` opts in to recognising an informal
vernacular ending (`"Coccinellid"` for family Coccinellidae) and trying an exact match on the formal
Latin equivalent before falling back to fuzzy matching.

## How well does it work?

The database's historical file records two things worth comparing separately: `taxon_name` (species
only) and `aligned_name` (every rank, in the exact `"Genus sp. [original; identifier]"` bracketed
format `include_bracketed_info = TRUE` reproduces).

```{r}
comparison <- old |>
  dplyr::select(original_name, taxon_name, old_aligned_name = aligned_name) |>
  dplyr::bind_cols(updated |> dplyr::select(new_aligned_name = aligned_name))
```

**Species-level (2,873 rows where `taxon_name` is filled in):** 2,861 (99.6%) match exactly. Only 12
disagree, and each one is informative rather than a package bug: six are species genuinely absent from
both AFD and iNaturalist (the pipeline conservatively stops at genus rather than guessing -- including
one, `"Euastacus coughrani"`, absent from both); two are likely fuzzy-matching imprecision within
*Euastacus*, a genus of dozens of very similarly-epithet-ed crayfish species, that reflects the limits of fuzzy matching, especially with incomplete resources and typos in submitted names; two are `"Onthophagus bulga"`
matched identically both times but recorded inconsistently in the *historical* file itself; one,
`"Incertae Sedis"`, resolves to AFD's own literal "position uncertain" placeholder rather than the more
specific subfamily a human curator assigned from context; and the last, `"Themognatha affinis
affinis"`, *was* a real bug -- a tautonymous subspecies textually indistinguishable from an affinity
hedge, fixed by checking for a real,
resource-verified exact match before ever falling back to the "uncertain identification" reading.

**Higher-rank (the other 2,281 rows):** matching the exact bracketed string is a stricter bar here --
format-sensitive in a way species names aren't -- and it shows: 1,911 (87.1%) match exactly. That's a
real drop from the species-level number, but the *same* two resource combinations agree with each
other on 98.5% of this subset, which suggests most of the gap is disagreement with the (older)
historical file, reflecting improvements in matching algorithms since this package was developed and the original code was upgraded.

## What real invertebrate names taught the matching engine

Several of `match_taxa()`'s ~20 sequential match blocks exist specifically because of patterns 
noted during test runs with the invertebrate dataset, but not anticipated in advance:

- **The bracketed subgenus convention** needed its own dedicated match blocks -- a plain string-cleaning
  pass (borrowed from APCalign, which never has to handle this convention) strips the parentheses but
  keeps the bracketed word, which can otherwise be mistaken for a species epithet and fuzzy-matched to
  an unrelated real species ([issue #14](https://github.com/traitecoevo/taxonAlign/issues/14)).
- **Fuzzy matching a higher rank needed to search broadest-first, not most-specific-first** -- checked
  directly against the real name list: 52% of names that fuzzy-matched a higher rank *also*
  fuzzy-matched a real candidate at a different rank, mostly not coincidence but a systematic pattern --
  an informal vernacular adjective (`"Coccinellid"`) must be assumed to signal the broadest group the name might refer to, not a
  specific genus ([issue #12](https://github.com/traitecoevo/taxonAlign/issues/12)), unless additional information is available.
- **Hybrid, intergrade, indecision, and graded/"affinis" identifications** all needed to resolve to
  genus rank only, since none of them can specify a genuine species -- and the "affinis" case
  specifically needed to check for a real exact match first, for the tautonym reason above
  ([issue #16](https://github.com/traitecoevo/taxonAlign/issues/16)).

The rules for aligning hybrids, intergrades, etc could be borrowed from APCalign's function, as they were already carefully calibrated. However, a plant taxonomist's reference
list essentially never contains a subgenus bracket or the informal references to higher rank groupings the way an invertebrate
collection's does.

## Swapping the reference: AFD + iNaturalist vs. AFD + GBIF

`taxonAlign` is designed to allow any taxonomic reference list to be used. To demonstrate this, the same 5,154-name list of invertebrate taxa was
aligned a second time, this time against AFD combined with a GBIF-backbone-derived reference of Australian
invertebrates, rather than reference lists from AFD + iNaturalist:

```{r}
gbif_au <- readRDS(system.file("extdata", "gbif_au_invertebrates.rds", package = "taxonAlign"))
resources_gbif <- prepare_taxonomic_resources(list(AFD = afd, GBIF = gbif_au))

updated_gbif <- create_taxonomic_update_lookup(
  old$original_name, resources_gbif,
  identifier = old$dataset_id,
  hybrids = TRUE, intergrades_affinis = TRUE, consider_english_name_endings = TRUE,
  include_bracketed_info = TRUE, progress = TRUE
)
```

The two combinations agree on 99.1% of all 5,154 names (98.5% on just the higher-rank subset, above).
Where they disagree, it's overwhelmingly explained by which resource happens to contain the specific
name -- good evidence that the reference and the matching engine really are two separable concerns, the
way the package's design intends.

## The GBIF chapter

`generate_GBIF_taxonomic_reference_list()` is the one piece of the package that doesn't need a
pre-existing export at all -- it builds a reference directly from the live GBIF backbone taxonomy for
any taxon:

```{r}
gbif_reference <- generate_GBIF_taxonomic_reference_list("Vanessa cardui")
```

Building a reference for *every* Australian invertebrate -- the scale a full trait database actually
needs -- was a substantially harder engineering problem than fetching one small clade:

- GBIF's taxonomy-browsing API refuses to page past 100,000 rows into any single clade -- a clade the
  size of Arthropoda (over three million descendants) has to be split recursively into smaller children
  and fetched piece by piece, with each piece cached independently as this became an oft-interrupted multi-day fetch.
- A country filter can't narrow the *fetch* itself -- GBIF's taxonomy-browsing endpoint has no country
  concept at all, only its occurrence-record endpoint does. The whole tree has to be fetched first and
  filtered to a country's occurrences afterward.
- Real network conditions at this scale -- transient TLS failures, requests that hang rather than error
  -- needed real retry-with-backoff and timeout handling, tuned against actual measured behaviour.

The result: a 358,496-row reference of Australian-occurring invertebrates, and a 3,987,219-row
worldwide version built the same way without the country filter -- confirming that a single worldwide
dataset built once could serve *any* country's alignment need, with the country filter applied locally
and cheaply afterward.

The plan is to eventually deliver the gbif taxonomic dataset together with a function that will allow it to be further filtered by taxon group or country. 

## Lessons for whoever reads this next

- **A hand-built test fixture finds the bugs you anticipated; a real, large, messy dataset finds the
  ones you didn't.** Several of the real bugs referenced above were only found by running the full
  name list end to end and diffing the result against independent alignments or a real historical
  answer key -- not by unit tests alone. The tests in `tests/testthat/` exist to *keep* each fixed once
  found, not to have found them in the first place. AI does a fantastic job of building tests to catch the anticipated bugs, greatly reducing the time costs of writing tests for a new package. A well-versed human curator is (currently still) required to dig deeper and deeper into the match sequence to identify additional bugs. A long list of messy taxon names is the only way these bugs are uncovered.
- **A plausible-sounding fix deserves the same scepticism as the bug it's fixing.** The first attempt (by AI!) at
  fixing issue #16 was itself wrong, and only caught because a real, adversarial test (initiated by a human)
  (`test-apc_equivalence.R`) existed to catch it.
- **A reference built for one purpose tends to be useful for another.** The GBIF reference built to
  validate this package against a real name list is, unmodified, also a candidate worldwide taxonomic
  resource in its own right.

See `CLAUDE.md` for the full technical detail behind every claim in this document, and the closed
issues on the [GitHub repository](https://github.com/traitecoevo/taxonAlign/issues?q=is%3Aissue) for
the discussion behind each fix.
