---
title: "Getting started with taxonAlign"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  warning: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`taxonAlign` helps with two related jobs:

1. **Build a personalised taxonomic reference** by combining one or more taxonomic reference tables --
   your own data, a list built from the GBIF backbone taxonomy, a known source like the Australian
   Faunal Directory (AFD) or the Australian Plant Census (APC), or any mix of these.
2. **Align and update a list of raw taxon names** against that reference -- matching misspellings,
   synonyms, and names only identifiable to a higher rank (`genus sp.`), then resolving each match
   forward to its current accepted name.

It's inspired by [APCalign](https://traitecoevo.github.io/APCalign/)'s approach to aligning taxon
names to the Australian Plant Census, generalised so the reference doesn't have to be APC/APNI and
the taxa don't have to be plants -- every function here works for any taxonomic rank present in your
reference, not just species. A real list of raw names is rarely all species anyway -- some
identifications only ever make it to genus, family, or a "morphospecies" code -- so every example
below deliberately mixes ranks, not just species names.

This vignette walks through every exported function, in the order you'd typically use them, then
puts them together into the workflow you'd actually run on a real list of names.

```{r setup}
library(taxonAlign)
library(dplyr)
```

## Building a taxonomic reference

Every matching/aligning function needs a `resources` object, built by `prepare_taxonomic_resources()`
from one or more taxonomic reference tables. There are three ways to get a reference table in the
first place.

### From your own data

If you already have a taxonomic reference table -- your own species list, an export from a museum
database, whatever -- `prepare_taxonomic_resources()` will take it as long as it has (or can be mapped
to) eight required columns: `canonical_name`, `scientific_name`, `taxon_rank`, `taxonomic_status`,
`taxonomic_dataset`, `genus`, `taxon_ID` and `accepted_name_usage_ID`.

Here's a small, hand-built example spanning several ranks: two genera, a family, a subgenus, a species
with two synonyms (one under a different genus entirely -- a real, common pattern when a species is
reclassified over its taxonomic history), and a subspecies.

```{r}
my_reference <- tibble::tribble(
  ~scientific_name,                          ~canonical_name,             ~taxon_rank,  ~taxonomic_status, ~taxonomic_dataset, ~genus,      ~taxon_ID, ~accepted_name_usage_ID,
  "Aporocera t-viride Blackburn, 1899",       "Aporocera t-viride",       "species",    "accepted",         "MY_DATA",          "Aporocera",  "sp1",    "sp1",
  "Aporocera viridipennis Lea, 1915",         "Aporocera viridipennis",   "species",    "synonym",          "MY_DATA",          "Aporocera",  "sp2",    "sp1",
  "Chalcolampra viridis Blackburn, 1897",     "Chalcolampra viridis",     "species",    "synonym",          "MY_DATA",          "Chalcolampra", "sp3",  "sp1",
  "Aporocera Blackburn",                      "Aporocera",                "genus",      "accepted",         "MY_DATA",          NA,           "g1",     "g1",
  "Aporocera (Aporocera) Blackburn",          "Aporocera",                "subgenus",   "accepted",         "MY_DATA",          "Aporocera",  "sg1",    "sg1",
  "Chrysomelidae Latreille",                  "Chrysomelidae",           "family",     "accepted",         "MY_DATA",          NA,           "f1",     "f1",
  "Xylotoles Pascoe",                          "Xylotoles",                "genus",      "accepted",         "MY_DATA",          NA,           "g2",     "g2",
  "Xylotoles costatus Pascoe, 1859",          "Xylotoles costatus",      "species",    "accepted",         "MY_DATA",          "Xylotoles",  "sp4",    "sp4",
  "Xylotoles costatus mixtus Someone, 1920",  "Xylotoles costatus mixtus", "subspecies", "accepted",       "MY_DATA",          "Xylotoles",  "sp5",    "sp5"
)

resources <- prepare_taxonomic_resources(my_reference)
names(resources)
```

`prepare_taxonomic_resources()` splits the combined table into a nested list -- one element per
taxonomic rank present (here: `species`, `genus`, `subgenus`, `family`), with `species` further split
into `accepted`/`synonym` sub-tables. This is the shape every other function in the package expects.

Note that `my_reference` above gives an explicit `"genus"`-rank row for both `"Aporocera"` and
`"Xylotoles"` -- but you don't actually have to. Naming a genus in the `genus` column of a species row
is enough on its own: `prepare_taxonomic_resources()` automatically adds a genus-rank row for any genus
(or any other rank you record in its own column, e.g. `family`) that's implied this way but doesn't yet
have an explicit row of its own, so `"Xylotoles sp. 1"` would still align to genus rank even if the
`"Xylotoles"` row above were deleted. Synthesised rows get a placeholder ID
(`taxonomic_dataset_taxon_rank_canonical_name`, e.g. `"MY_DATA_genus_Xylotoles"`) since there's no
natural ID to reuse.

If your table doesn't have these columns under these exact names, pass `interactive = TRUE` and you'll
be prompted (in the style of `traits.build`'s `metadata_add_traits()`) to map your own column names, or
supply a fixed value, instead of erroring. If you don't even have a table loaded yet, you can omit
`taxonomic_resources` entirely and you'll be prompted for a file path to start from:

```{r}
#| eval: false
prepare_taxonomic_resources(interactive = TRUE)
#> Enter the file path for your taxonomic reference: my_beetles.csv
#> Does the data have a column for `taxon_rank`, or is every row the same value? ...
```

See `?prepare_taxonomic_resources` for the full set of prompts.

### From the GBIF backbone taxonomy

If you don't have your own reference, `generate_GBIF_taxonomic_reference_list()` builds one from the
GBIF backbone taxonomy for any taxon name, optionally restricted to a country and/or a minimum rank.
The whole subtree is fetched once and cached, so repeat calls are fast.

```{r}
gbif_resources <- generate_GBIF_taxonomic_reference_list("Aporocera", country = "AU") |>
  prepare_taxonomic_resources()
```

Real GBIF data can include a handful of "doubtful" or incompletely-classified records with no
canonical name at all -- `prepare_taxonomic_resources()` drops these automatically (with a warning),
since a name with no usable label could never be matched against anyway.

### From a known source (AFD, APC)

`load_taxonomic_resources()` fetches and reshapes a fixed set of *known* datasets into the same flat
schema -- currently the Australian Faunal Directory (`"AFD"`) and a wrapper around
`APCalign::load_taxonomic_resources()` (`"APC"`). Unlike the GBIF function, these aren't built from a
taxon name -- they're a whole dataset at once:

```{r}
#| eval: false
afd <- load_taxonomic_resources("AFD")   # reads/reshapes a local AFD export (see ?load_taxonomic_resources)
apc <- load_taxonomic_resources("APC")   # wraps APCalign::load_taxonomic_resources()
```

`load_taxonomic_resources()` always returns a named list, even for one dataset, so both pieces above
can be combined directly.

### Combining multiple reference tables

`prepare_taxonomic_resources()` also accepts a list of tables, combining them with a priority order:
when the same name appears in more than one table, whichever table is listed *first* wins. This is how
you'd combine a known source with your own data, or two known sources together, when you already have
each table in hand:

```{r}
#| eval: false
resources <- prepare_taxonomic_resources(list(mine = my_reference, gbif = gbif_resources))
```

If you'd rather be walked through it -- especially useful when you're adding sources one at a time
rather than assembling them all up front -- `interactive = TRUE` handles the whole thing itself: it
asks for the first table (a path, exactly as above), walks through its column mapping, then asks "Do
you have any additional taxonomic reference(s) to include?" -- repeating for as many as you have. You
never construct the combined list yourself; the function builds it internally as you go.

```{r}
#| eval: false
prepare_taxonomic_resources(interactive = TRUE)
#> Enter the file path for your taxonomic reference: my_beetles.csv
#> ... (column mapping for my_beetles.csv) ...
#> Do you have any additional taxonomic reference(s) to include? 1: Yes 2: No
#> Selection: 1
#> Enter the file path for the additional taxonomic reference: more_beetles.csv
#> ... (column mapping for more_beetles.csv) ...
#> Do you have any additional taxonomic reference(s) to include? 1: Yes 2: No
#> Selection: 2
#> Multiple taxonomic reference tables were supplied. Enter their priority order ...
```

## Matching raw names: `align_taxa()`

`align_taxa()` takes one or more raw taxon names and matches each against `resources`, trying (in
order) exact matches, then progressively more lenient fuzzy matches, at species level first and then
at each higher rank present in `resources`.

```{r}
align_taxa("Aporocera viridipennis", resources)  # exact match, a synonym
align_taxa("Aporocera viridipenis", resources)   # one letter missing -- fuzzy match
align_taxa("Xylotoles costatus mixtus", resources) # a subspecies
align_taxa("Chrysomelidae", resources)            # family only
```

A name that can't be resolved to anything is left `NA`, not an error:

```{r}
align_taxa("Completely unrelated nonsense", resources)
```

### Matching more than one name at once

`align_taxa()` is vectorised -- pass as many names as you like in one call, spanning as many different
ranks as your raw data actually has:

```{r}
align_taxa(
  c("Aporocera t-viride", "Aporocera viridipenis", "Xylotoles costatus mixtus", "Chrysomelidae", "Xylotoles sp. 1"),
  resources
) |>
  select(original_name, aligned_name, taxon_rank)
```

In practice your names usually live in a column of a data frame, not a bare vector -- just pass that
column straight in:

```{r}
field_data <- tibble(
  name = c("Aporocera t-viride", "Chalcolampra viridis", "Xylotoles sp. 1", "Totally unknown beetle"),
  site = c("Site A", "Site B", "Site C", "Site B")
)

align_taxa(field_data$name, resources) |>
  select(original_name, aligned_name, taxon_rank, taxonomic_status)
```

Two opt-in parameters handle names that structurally can't specify a species at all, resolving them to
genus rank instead of leaving them to (possibly mis-)match a species-level pattern:

```{r}
align_taxa("Aporocera x hybridname", resources, hybrids = TRUE)
align_taxa("Aporocera aff. viridipennis", resources, intergrades_affinis = TRUE)
```

`hybrids` catches `Genus x species`-style hybrid names; `intergrades_affinis` catches intergrades
(`--`), a collector's indecision between two taxa (`/`), and graded/uncertain identifications
(`aff.`/`cf.`). Both default to `FALSE`.

For a large list, `progress = TRUE` prints a text progress bar tracking how many names have been
resolved so far.

## Resolving to current names: `update_taxa()`

`align_taxa()` tells you *what a name matched* -- `update_taxa()` takes that result and resolves it
forward to whatever the *current* accepted name is, following synonym links via
`accepted_name_usage_ID`. This works the same way regardless of rank -- a species-level synonym and a
higher-rank one both resolve via the same lookup:

```{r}
aligned <- align_taxa(c("Chalcolampra viridis", "Xylotoles sp. 1"), resources)
update_taxa(aligned, resources) |>
  select(original_name, aligned_name, accepted_name, taxonomic_status)
```

`accepted_name` is the current accepted name (`NA` if it couldn't be resolved); `suggested_name` falls
back to `aligned_name` when there's no `accepted_name`, so it's never `NA` for anything that matched at
all.

## The one-call pipeline: `create_taxonomic_update_lookup()`

Most of the time you want both steps together. `create_taxonomic_update_lookup()` runs `align_taxa()`
then `update_taxa()` in one call:

```{r}
create_taxonomic_update_lookup(
  c("Aporocera viridipennis", "Aporocera viridipenis", "Chalcolampra viridis", "Xylotoles costatus mixtus", "Chrysomelidae"),
  resources
) |>
  select(original_name, aligned_name, accepted_name, taxonomic_status, update_reason)
```

## Putting it together: updating a list of names

Here's the workflow you'd actually run on a real list of raw names -- say, taxon names recorded in a
field dataset that need reconciling against a reference before analysis, spanning several ranks and
including a couple of names nothing will match.

```{r}
field_data <- tibble(
  original_name = c(
    "Aporocera viridipennis",   # a synonym
    "Aporocera viridipenis",    # typo
    "Xylotoles costatus mixtus", # a subspecies
    "Chrysomelidae",             # family only
    "Aporocera sp. 1",          # morphospecies code -- genus known, species not
    "Totally unknown beetle"    # won't resolve to anything
  ),
  site = c("Site A", "Site B", "Site A", "Site C", "Site B", "Site B")
)

lookup <- create_taxonomic_update_lookup(field_data$original_name, resources, identifier = field_data$site)
lookup |> select(original_name, identifier, suggested_name, taxonomic_status, update_reason)
```

`identifier` (here, `site`) is threaded through so that a name only ever resolvable to genus rank (like
the morphospecies code above) is disambiguated per dataset/location in its aligned name, rather than
collapsing two genuinely different unidentified morphospecies from different sites into one.

Join the lookup back onto your original data by `original_name` to get an updated name for every row:

```{r}
field_data |> left_join(lookup, by = "original_name")
```

From here, the rows worth a closer look are the ones where `taxonomic_status` is `"unknown"` (nothing
matched at all) or `suggested_name` is `NA` -- everything else resolved to *something*, even if only at
a higher rank. Reviewing exactly those rows (rather than the whole list) is usually the fastest way to
find the handful of names that need a manual look -- a real typo not caught by fuzzy matching, a name
missing from your reference entirely, or a genuinely new/undescribed taxon.

```{r}
lookup |> filter(is.na(suggested_name) | taxonomic_status == "unknown")
```
