---
title: "Getting started with taxonAlign"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  warning: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`taxonAlign` helps with two related jobs:

1. **Build a personalised taxonomic reference** by combining one or more taxonomic reference
   tables -- your own data, a list built from the GBIF backbone taxonomy, a known source like the
   Australian Faunal Directory (AFD) or the Australian Plant Census (APC), or any mix of these.
2. **Align and update a list of raw taxon names** against that reference -- matching misspellings,
   synonyms, and names only identifiable to a higher rank (`genus sp.`), then resolving each match
   forward to its current accepted name.

It's inspired by [APCalign](https://traitecoevo.github.io/APCalign/)'s approach to aligning taxon
names to the Australian Plant Census, generalised so the reference doesn't have to be APC/APNI and
the taxa don't have to be plants -- every function here works for any taxonomic rank present in your
reference, not just genus/species/family.

This vignette walks through every exported function, in the order you'd typically use them, then
puts them together into the workflow you'd actually run on a real list of names.

```{r setup}
library(taxonAlign)
library(dplyr)
```

## Building a taxonomic reference

Every matching/aligning function needs a `resources` object, built by `prepare_taxonomic_resources()`
from one or more taxonomic reference tables. There are three ways to get a reference table in the
first place.

### From your own data

If you already have a taxonomic reference table -- your own species list, an export from a museum
database, whatever -- `prepare_taxonomic_resources()` will take it as long as it has (or can be
mapped to) eight required columns: `canonical_name`, `scientific_name`, `taxon_rank`,
`taxonomic_status`, `taxonomic_dataset`, `genus`, `taxon_ID` and `accepted_name_usage_ID`.

Here's a small, hand-built example spanning a genus, two species (one a synonym of the other) and a
family:

```{r}
my_reference <- tibble::tribble(
  ~scientific_name,                    ~canonical_name,          ~taxon_rank, ~taxonomic_status, ~taxonomic_dataset, ~genus,      ~taxon_ID, ~accepted_name_usage_ID,
  "Aporocera viridipennis Lea, 1915",   "Aporocera viridipennis", "species",   "accepted",         "MY_DATA",          "Aporocera", "sp1",     "sp1",
  "Chalcolampra viridis Blackburn, 1897", "Chalcolampra viridis", "species",   "synonym",          "MY_DATA",          "Chalcolampra", "sp2",  "sp1",
  "Aporocera Blackburn",                "Aporocera",             "genus",     "accepted",         "MY_DATA",          NA,          "g1",      "g1",
  "Chrysomelidae Latreille",            "Chrysomelidae",         "family",    "accepted",         "MY_DATA",          NA,          "f1",      "f1"
)

resources <- prepare_taxonomic_resources(my_reference)
names(resources)
```

`prepare_taxonomic_resources()` splits the combined table into a nested list -- one element per
taxonomic rank present (here: `species`, `genus`, `family`), with `species` further split into
`accepted`/`synonym` sub-tables. This is the shape every other function in the package expects.

If your table doesn't have these columns under these exact names, pass `interactive = TRUE` and
you'll be prompted (in the style of `traits.build`'s `metadata_add_traits()`) to map your own column
names, or supply a fixed value, instead of erroring -- see `?prepare_taxonomic_resources` for details.

### From the GBIF backbone taxonomy

If you don't have your own reference, `generate_GBIF_taxonomic_reference_list()` builds one from the
GBIF backbone taxonomy for any taxon name, optionally restricted to a country and/or a minimum rank.
The whole subtree is fetched once and cached, so repeat calls are fast.

```{r}
gbif_reference <- generate_GBIF_taxonomic_reference_list("Aporocera", country = "AU")
gbif_reference |> select(canonical_name, taxon_rank, taxonomic_status)
```

### From a known source (AFD, APC)

`load_taxonomic_resources()` fetches and reshapes a fixed set of *known* datasets into the same flat
schema -- currently the Australian Faunal Directory (`"AFD"`) and a wrapper around
`APCalign::load_taxonomic_resources()` (`"APC"`). Unlike the GBIF function, these aren't built from a
taxon name -- they're a whole dataset at once:

```{r}
#| eval: false
afd <- load_taxonomic_resources("AFD")   # reads/reshapes a local AFD export (see ?load_taxonomic_resources)
apc <- load_taxonomic_resources("APC")   # wraps APCalign::load_taxonomic_resources()
```

`load_taxonomic_resources()` always returns a named list, even for one dataset, so both pieces above
can be combined directly.

### Combining multiple reference tables

`prepare_taxonomic_resources()` also accepts a list of tables, combining them with a priority order:
when the same name appears in more than one table, whichever table is listed *first* wins. This is
how you'd combine a known source with your own data, or two known sources together:

```{r}
#| eval: false
resources <- prepare_taxonomic_resources(
  list(AFD = load_taxonomic_resources("AFD")$AFD, mine = my_reference)
)
```

## Matching raw names: `align_taxa()`

`align_taxa()` takes a vector of raw taxon names and matches each one against `resources`, trying
(in order) exact matches, then progressively more lenient fuzzy matches, at species level first and
then at each higher rank present in `resources`.

```{r}
align_taxa("Aporocera viridipennis", resources)  # exact match
align_taxa("Aporocera viridipenis", resources)   # one letter missing -- fuzzy match
align_taxa("Chalcolampra viridis", resources)    # a synonym, correctly flagged as such
align_taxa("Aporocera", resources)                # genus only
```

A name that can't be resolved to anything is left `NA`, not an error:

```{r}
align_taxa("Completely unrelated nonsense", resources)
```

Two opt-in parameters handle names that structurally can't specify a species at all, resolving them
to genus rank instead of leaving them to (possibly mis-)match a species-level pattern:

```{r}
align_taxa("Aporocera x hybridname", resources, hybrids = TRUE)
align_taxa("Aporocera aff. viridipennis", resources, intergrades_affinis = TRUE)
```

`hybrids` catches `Genus x species`-style hybrid names; `intergrades_affinis` catches intergrades
(`--`), a collector's indecision between two taxa (`/`), and graded/uncertain identifications
(`aff.`/`cf.`). Both default to `FALSE`.

For a large list, `progress = TRUE` prints a text progress bar tracking how many names have been
resolved so far.

## Resolving to current names: `update_taxa()`

`align_taxa()` tells you *what a name matched* -- `update_taxa()` takes that result and resolves it
forward to whatever the *current* accepted name is, following synonym links via
`accepted_name_usage_ID`:

```{r}
aligned <- align_taxa("Chalcolampra viridis", resources)
update_taxa(aligned, resources) |> select(original_name, aligned_name, accepted_name, taxonomic_status)
```

`accepted_name` is the current accepted name (`NA` if it couldn't be resolved); `suggested_name` falls
back to `aligned_name` when there's no `accepted_name`, so it's never `NA` for anything that matched
at all.

## The one-call pipeline: `create_taxonomic_update_lookup()`

Most of the time you want both steps together. `create_taxonomic_update_lookup()` runs `align_taxa()`
then `update_taxa()` in one call:

```{r}
create_taxonomic_update_lookup(
  c("Aporocera viridipennis", "Aporocera viridipenis", "Chalcolampra viridis", "Aporocera"),
  resources
) |> select(original_name, aligned_name, accepted_name, taxonomic_status, update_reason)
```

## Putting it together: updating a list of names

Here's the workflow you'd actually run on a real list of raw names -- say, taxon names recorded in a
field dataset that need reconciling against a reference before analysis.

```{r}
field_data <- tibble(
  original_name = c(
    "Aporocera viridipennis",   # exact match
    "Aporocera viridipenis",    # typo
    "Chalcolampra viridis",     # synonym
    "Aporocera sp. 1",          # morphospecies code -- genus known, species not
    "Totally unknown beetle"    # won't resolve to anything
  ),
  site = c("Site A", "Site B", "Site A", "Site C", "Site B")
)

lookup <- create_taxonomic_update_lookup(field_data$original_name, resources, identifier = field_data$site)
lookup |> select(original_name, identifier, suggested_name, taxonomic_status, update_reason)
```

`identifier` (here, `site`) is threaded through so that a name only ever resolvable to genus rank
(like the morphospecies code above) is disambiguated per dataset/location in its aligned name, rather
than collapsing two genuinely different unidentified morphospecies from different sites into one.

Join the lookup back onto your original data by `original_name` to get an updated name for every row:

```{r}
field_data |> left_join(lookup, by = "original_name")
```

From here, the rows worth a closer look are the ones where `taxonomic_status` is `"unknown"` (nothing
matched at all) or `suggested_name` is `NA` -- everything else resolved to *something*, even if only
at genus rank. Reviewing exactly those rows (rather than the whole list) is usually the fastest way to
find the handful of names that need a manual look -- a real typo not caught by fuzzy matching, a name
missing from your reference entirely, or a genuinely new/undescribed taxon.

```{r}
lookup |> filter(is.na(suggested_name) | taxonomic_status == "unknown")
```
