---
title: "reproduce-EH-workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{reproduce-EH-workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(taxonAlign)
library(tidyverse)
library(here)
```

## User to load taxonomic resources

```{r}
taxon_resources <- read_csv("../ausinvertraits.addons/data_extra/taxon_list_for_AusInvertAlign.csv")

# Preview
taxon_resources
names(taxon_resources)
```

### Prepare and split taxonomic resources

`prepare_taxonomic_resources()` adds the extra columns the matching engine needs (stripped/cleaned
name variants, binomial/trinomial, etc.) and splits the combined table into the nested `resources`
list `align_taxa()` expects -- by taxon rank, further splitting `species` by taxonomic status, and
building both subgenus-matching conventions (see `?prepare_taxonomic_resources`).

```{r}
resources <- prepare_taxonomic_resources(taxon_resources)

names(resources)
names(resources$species)
names(resources$subgenus_v2)
```

### Use AusInvertTraits Database a user reference list to check with the taxonomic resource

```{r}
# I can't seem to be able to remake ausinverttraits so I am reverse engineering the user's name list with the actual "answer" output
ausinvertraits_aligned_names <- read_csv(here("../ausinvertraits.addons/data_extra/aligned_names.csv"))

ausinvertraits_aligned_names$original_name |> unique() |> length()

ausinvertraits_names 
```

```{r}
taxa_to_check <- ausinvertraits_aligned_names |> distinct(original_name, identifier)
```

## Align taxa

```{r}
# Takes a few minutes this one
aligned_names_b <- align_taxa(
    original_name = taxa_to_check$original_name,
    resources = resources,
    identifier = taxa_to_check$identifier,
    fuzzy_abs_dist = 3,
    fuzzy_rel_dist = 0.2,
    fuzzy_matches = TRUE,
    imprecise_fuzzy_matches = FALSE
)

saveRDS(aligned_names_b, "data/aligned_names_b.rds")

# Not matching here
nrow(aligned_names_b)
nrow(ausinvertraits_aligned_names)

# So what's different?
dim(aligned_names_b)
dim(ausinvertraits_aligned_names)

# Okay just in number of rows - these must be in the ausinverttraits database prior to the creation of ausinvertraits_aligned_names
rows_diff <- setdiff(ausinvertraits_aligned_names$original_name, aligned_names_b$original_name)

ausinvertraits_aligned_names |>
  filter(original_name %in% rows_diff) |>
  select(original_name, aligned_name) |>
  count(aligned_name)
```

